Game theory and language: a cautionary tale

### 2025-06-17 manuscript/manuscript.Rmd
many graph much fuss

---
title: "Game theory and language: a cautionary tale"
authors:
  - name: Veronica Boyce
    department: Department of Psychology
    affiliation: Stanford University
    email: vboyce@stanford.edu
  - name: Avidit Acharya
    department: Department of Political Science
    affiliation: Stanford University
  - name: Michael C. Frank
    department: Department of Psychology
    affiliation: Stanford University
abstract: |
  An attempt was made TODO
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    extra_dependencies: ["float"]

@@

As shown in Figure TODO, groups vary substantially in how many trials they use the chat on, but many groups rarely use the chat. The horizontal lines correspond to 10% and 50% of the time, so games lower than the 10% line are using chat on fewer than 4 (out of 40) trials. There's a lot of not chatting in the access-to-chat condition.

```{r chat-pct, fig.width=10, fig.height=3, }

@@

game_chat |>
  filter(chat_cond == "chat") |>
  mutate(is.chat = ifelse(words > 0, 1, 0)) %>%
  group_by(gameId, expt) %>%
  summarize(pct_chat = mean(is.chat)) |>
  arrange(expt, pct_chat) |>
  group_by(expt) |>
  mutate(x = 1 / n() * row_number()) |>
  ggplot(aes(x = x, y = pct_chat)) +
  geom_point() +
  facet_grid(.~expt) +
  coord_cartesian(ylim = c(-.05, 1.05),xlim=c(-.05, 1.05), expand=F) +
  geom_hline(yintercept = .1) +
  geom_hline(yintercept = .5) +
  labs(x="Fraction games", y="Fraction trials with chat messages")
```

@@

@@

The first way we assess whether access to chat makes a difference on game performance is to look at the mean number of points recieved per trial. Because BoS and PD have different point structures, we compare within game type.

```{r points_pd, fig.cap="TODO fig cap"}
bonuses <- all_rounds %>%
  group_by(playerId, gameId, chat_cond, cond, gametype, expt) %>%
  summarize(payoff = mean(payoff))

ggplot(bonuses, aes(x = chat_cond, y = payoff, color = str_c(cond, "_", gametype), group = cond)) +
  geom_point(position = position_jitterdodge(jitter.width = .1, dodge.width = .7), alpha = .3) +
  facet_grid(gametype~expt) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "mean_reward", x = "condition") +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .7), size = .7, color = "black") +
    stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .7), size = .7, color = "black", geom="line") +

  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0, 20)) +
  labs(y = "Mean reward / trial", ) +
  theme(
    legend.title=element_blank(),
    axis.title.x=element_blank(),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14)
  )
```

TODO set per-condition color scheme!

@@

TODO include topline model results

@@

```{r}
outcome <- all_rounds %>%
  select(gameId, repNum, response, chat_cond, gametype, cond, role) |>
  pivot_wider(names_from = role, values_from = response) %>%
  mutate(outcome = str_c(p1, p2)) %>%
  left_join(game_chat)
```

For BoS, the desirable outcome is to both choose the same box. As can be seen in Figure TODO, this good outcome is achieved at above change levels when people can chat and usually at chance levels when people cannot.

```{r, fig.cap="In BoS: P1 prefers AA to BB, P2 prefers BB to AA. AB and BA are 0 for both. "}
outcome %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond), y = outcome_parity, color=cond, group=cond)) +
  facet_grid(~ expt) +
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  labs(title = "BoS:fraction choosing good outcomes (AA or BB)", y = "Good outcome?")+
  theme(legend.position="bottom")
```
We again note that the off-diagonal rewards were misdisplayed for one of the players in experiments 1 and 2 for PD. On most trials "normal_PD" will pattern with easy PD.

@@

```{r}
outcome %>%
  filter(gametype == "PD") %>%
  mutate(AA = ifelse(outcome %in% c("AA"), 1, 0)) %>%
  mutate(AB_BA= ifelse(outcome %in% c("AB", "BA"), 1, 0)) %>%
  mutate(BB = ifelse(outcome %in% c("BB"), 1, 0)) %>%
  pivot_longer(c("AA", "AB_BA", "BB"), names_to="outcome_type", values_to="selected") |> 
  ggplot(aes(x = chat_cond, y = selected, color=cond)) +
  facet_grid(outcome_type ~ expt) +
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  theme(legend.position="bottom")
```
From Figure TODO, we see that the preferred outcome is AA which is welfare-maximizing in easy PD. This is not what game theory predicts for rational actors, instead, participants are being nice and cooperative with each other.

There is a slight tendency for more AA selections in easy than hard PD, especially in the chat conditions, but slight.

# Analyses based on use of chat

## Per game does talking help?

One idea is that talking on some trials may set up coordination strategies that can then effectively be used on later trials without talking on those trials.

So we want to look at overall volume of talking (in words or in # of trial talked) as a predictor for performance, controlling for talk on that round?

```{r}
group_talk <- game_chat |>
  group_by(gameId) |>
  summarize(total_words = sum(words), pct_chat = mean(is_chat))

overall_talk <- game_chat |> left_join(group_talk)
ggplot(overall_talk |> filter(gametype == "BoS", chat_cond == "chat"), aes(x = pct_chat, y = mean_payout, color = as.character(is_chat))) +
  # stat_summary(fun.data="mean_cl_boot")+
  # geom_point()+
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")

ggplot(overall_talk |> filter(gametype == "PD", chat_cond == "chat"), aes(x = pct_chat, y = mean_payout, color = as.character(is_chat))) +
  # stat_summary(fun.data="mean_cl_boot")+
  # geom_point()+
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")
```

So, in mixed games, talking on more trials helps regardess of whether you talked on this particular trial. (Although overall talk might also be confounding with conscientiousness..., to fix you'd have to experimentally manipulate when chat is or isn't available)

Talking more might help with hard PD, maybe in mixed?

```{r}
ggplot(overall_talk |> filter(gametype == "BoS", total_words > 0), aes(x = log(total_words), y = mean_payout, color = as.character(is_chat))) +
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")

ggplot(overall_talk |> filter(gametype == "PD", total_words > 0), aes(x = log(total_words), y = mean_payout, color = as.character(is_chat))) +
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")
```

Here total words is going to mix how much you talk each time with how often you talk, so maybe this is just totally redundant with previous. More talk helps on BoS, even as transfer from other trials. PD seems messier?

## Quadrant analysis per trial

 - chat_0 is had access to chat and didn't use on that trial
 - chat_1 is used chat on that trial
 - nochat_0 did not have access to chat

```{r}
outcome <- all_rounds %>%
  select(gameId, repNum, response, chat_cond, gametype, cond, role) |>
  pivot_wider(names_from = role, values_from = response) %>%
  mutate(outcome = str_c(p1, p2)) %>%
  left_join(game_chat)
```

## BoS

@@

```{r}
outcome %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond, "_", is_chat), y = outcome_parity)) +
  facet_grid(. ~ cond) +
  stat_summary(fun.data = "mean_cl_boot") +
  geom_hline(aes(yintercept = .5)) +
  labs(title = "BoS:fraction choosing good outcomes (AA or BB)", y = "Good outcome?")
```

Near chance if you can't talk, above chance if you could but don't,  far above chance if you do.

```{r}
outcome %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = repNum, y = outcome_parity, color = cond)) +
  facet_grid(~ str_c(chat_cond, "_", is_chat)) +
  geom_smooth(method = "lm", se = F) +
  geom_hline(aes(yintercept = .5)) +
  labs(title = "BoS:fraction choosing good outcomes", y = "Good outcome?")
```

## PD

Expts 1 and 2 are suspect, but on most trials will pattern with easy PD.

```{r}
outcome %>%
  filter(gametype == "PD") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond, "_", is_chat), y = outcome_parity)) +
  facet_grid(. ~ cond) +
  stat_summary(fun.data = "mean_cl_boot") +
  geom_hline(aes(yintercept = .5)) +
  labs(title = "PD:fraction choosing AA outcome")

outcome %>%
  filter(gametype == "PD") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AB", "BA"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond, "_", is_chat), y = outcome_parity)) +
  facet_grid(. ~ cond) +
  stat_summary(fun.data = "mean_cl_boot") +
  geom_hline(aes(yintercept = .5)) +
  labs(title = "PD:fraction choosing AB or BA outcome")

outcome %>%
  filter(gametype == "PD") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("BB"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond, "_", is_chat), y = outcome_parity)) +
  facet_grid(. ~ cond) +
  stat_summary(fun.data = "mean_cl_boot") +
  geom_hline(aes(yintercept = .5)) +
  labs(title = "PD:fraction choosing BB outcome")
```

Can get a reasonable option no matter what. If you talk, you can get the uneven, but welfare-maximizing in hard PD.

Looks like using language tends to help, on the trials it's used on, and slightly on the trials it's not used on for BoS. Helps in some cases for PD.

Is there a dose-response relationship, or is one word enough?

```{r}
ggplot(game_chat |> filter(gametype == "BoS", words > 0), aes(x = log(words), y = mean_payout)) +
  geom_point() +
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")

ggplot(game_chat |> filter(gametype == "PD", words > 0), aes(x = log(words), y = mean_payout)) +
  geom_point() +
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")
```

@@

## Per game does talking help?

One idea is that talking on some trials may set up coordination strategies that can then effectively be used on later trials without talking on those trials.

So we want to look at overall volume of talking (in words or in # of trial talked) as a predictor for performance, controlling for talk on that round?

```{r}
group_talk <- game_chat |>
  group_by(gameId) |>
  summarize(total_words = sum(words), pct_chat = mean(is_chat))

overall_talk <- game_chat |> left_join(group_talk)
ggplot(overall_talk |> filter(gametype == "BoS", chat_cond == "chat"), aes(x = pct_chat, y = mean_payout, color = as.character(is_chat))) +
  # stat_summary(fun.data="mean_cl_boot")+
  # geom_point()+
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")

ggplot(overall_talk |> filter(gametype == "PD", chat_cond == "chat"), aes(x = pct_chat, y = mean_payout, color = as.character(is_chat))) +
  # stat_summary(fun.data="mean_cl_boot")+
  # geom_point()+
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")
```

So, in mixed games, talking on more trials helps regardess of whether you talked on this particular trial. (Although overall talk might also be confounding with conscientiousness..., to fix you'd have to experimentally manipulate when chat is or isn't available)

Talking more might help with hard PD, maybe in mixed?

```{r}
ggplot(overall_talk |> filter(gametype == "BoS", total_words > 0), aes(x = log(total_words), y = mean_payout, color = as.character(is_chat))) +
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")

ggplot(overall_talk |> filter(gametype == "PD", total_words > 0), aes(x = log(total_words), y = mean_payout, color = as.character(is_chat))) +
  facet_grid(gametype ~ cond) +
  geom_smooth(method = "lm")
```

Here total words is going to mix how much you talk each time with how often you talk, so maybe this is just totally redundant with previous. More talk helps on BoS, even as transfer from other trials. PD seems messier?

# Language results: what do people say
and what does this say about coordination strategies

## Language

How much language?

Filter only for games that talked at least a little.

Second graph filters for *trials* that talked.

```{r}
game_chat |>
  group_by(gameId) |>
  summarize(total_words = sum(words)) |>
  filter(total_words != 0) |>
  left_join(game_chat) |>
  mutate(expt = str_sub(cond, -1)) |>
  ggplot(aes(x = repNum, y = words)) +
  geom_point() +
  facet_wrap(~expt) +
  geom_smooth()

game_chat |>
  group_by(gameId) |>
  summarize(total_words = sum(words)) |>
  filter(total_words != 0) |>
  left_join(game_chat) |>
  filter(words > 0) |>
  mutate(expt = str_sub(cond, -1)) |>
  ggplot(aes(x = repNum, y = words)) +
  geom_point() +
  facet_wrap(~expt) +
  geom_smooth()
```

## how often do *both* people talk

```{r}
pair_chat <- all_chat %>%
  filter(!is.na(targets)) %>%
  select(-type) %>%
  full_join(all_rounds) %>%
  mutate(words = str_count(text, "\\W+") %|% int(0)) |>
  group_by(gameId, playerId, gametype, cond, chat_cond, repNum) %>%
  summarize(
    words = sum(words),
    is_chat = ifelse(words > 0, 1, 0)
  ) |>
  group_by(gameId, gametype, cond, chat_cond, repNum) |>
  summarize(how_many_chat = sum(is_chat)) |>
  group_by(gameId, gametype, cond, chat_cond) |>
  summarize(pct_both = mean(how_many_chat == 2)) |>
  filter(chat_cond == "chat")

ggplot(pair_chat, aes(x = cond, y = pct_both)) +
  geom_jitter(width = .2) +
  facet_wrap(~gametype)
```
Even in games that talk, there aren't *that* many trials where both people talk?

Probably going to need to manually look at the language for negotiation v "choose red" "yup" etc.

@@

3.

@@

```{r}
do_table("payout_mega.rds", cap="TODO")
do_table("payout_mega_combo.rds", cap="TODO")
do_table("payout_combo_3.rds", cap="TODO")
do_table("payout_combo_4.rds", cap="TODO")
do_table("payout_combo_5.rds", cap="TODO")

do_table("BoS_aligned_mega.rds", cap="TODO")
do_table("BoS_aligned_3.rds", cap="TODO")
do_table("BoS_aligned_4.rds", cap="TODO")
do_table("BoS_aligned_5.rds", cap="TODO")

do_table("PD_cooperate_mega.rds", cap="TODO")
do_table("PD_cooperate_3.rds", cap="TODO")
do_table("PD_cooperate_4.rds", cap="TODO")
do_table("PD_cooperate_5.rds", cap="TODO")

do_table("PD_unneven_mega.rds", cap="TODO")
do_table("PD_unneven_3.rds", cap="TODO")
do_table("PD_unneven_4.rds", cap="TODO")
do_table("PD_unneven_5.rds", cap="TODO")

do_table("PD_nash_mega.rds", cap="TODO")
do_table("PD_nash_3.rds", cap="TODO")
do_table("PD_nash_4.rds", cap="TODO")
do_table("PD_nash_5.rds", cap="TODO")

@@

 [1] "BoS_aligned_3.rds"     "BoS_aligned_4.rds"     "BoS_aligned_5.rds"    
 [4] "BoS_aligned_mega.rds"  "payout_mega_3.rds"     "payout_mega_4.rds"    
 [7] "payout_mega_combo.rds" "payout_mega.rds"       "PD_cooperate_3.rds"   
[10] "PD_cooperate_4.rds"    "PD_cooperate_5.rds"    "PD_cooperate_mega.rds"
[13] "PD_nash_3.rds"         "PD_nash_4.rds"         "PD_nash_5.rds"        
[16] "PD_nash_mega.rds"      "PD_unneven_3.rds"      "PD_unneven_4.rds"     
[19] "PD_unneven_5.rds"      "PD_unneven_mega.rds"

## payout models

First up, payout models.

## BoS alignment models

## PD alignment models

TODO run Bayesian
TODO save Bayesian
TODO tables

# Random crap

## Condition assignment on performance

```{r, eval=F}
perf_priors <- c(
  set_prior("normal(5,2)", class = "Intercept"),
  set_prior("normal(0, 2)", class = "b"),
  set_prior("normal(0, 2)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

perf_mod <- brm(
  payoff ~ has_chat * is_BoS + repNum +
    (is_BoS | gameId) +
    (is_BoS | playerId),
  data = for_mod,
  prior = perf_priors,
  control = list(adapt_delta = .95)
)

perf_mod_tot <- brm(
  payoff ~ (use_chat | has_chat) * is_BoS +
    repNum +
    (is_BoS | gameId) +
    (is_BoS | playerId),
  data = for_mod,
  prior = perf_priors,
  control = list(adapt_delta = .95)
)

freq_tot <- lm(payoff ~ (use_chat | has_chat) * is_BoS +
  repNum, data = for_mod)
```

## Words over time

log(words) ~ game_type * trial_number

```{r, eval=F}
for_log_words <- for_mod |>
  filter(words > 0) |>
  mutate(log_words = log(words))

lang_priors <- c(
  set_prior("normal(5,5)", class = "Intercept"),
  set_prior("normal(0, 5)", class = "b"),
  set_prior("normal(0, 5)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

log_words_mod <- brm(
  log_words ~ is_BoS * repNum +
    (is_BoS * repNum | gameId),
  data = for_log_words,
  prior = lang_priors,
  control = list(adapt_delta = .95)
)
```

## Words on outcome

outcome ~ game_type * log(words)

>>>>

---
title: "Game theory and language: a cautionary tale"
authors:
  - name: Veronica Boyce
    department: Department of Psychology
    affiliation: Stanford University
    email: vboyce@stanford.edu
  - name: Avidit Acharya
    department: Department of Political Science
    affiliation: Stanford University
  - name: Michael C. Frank
    department: Department of Psychology
    affiliation: Stanford University
abstract: |
  An attempt was made TODO
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    extra_dependencies: ["float"]
header-includes:
 - \usepackage{float}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{5}
 - \setcounter{bottomnumber}{5}
 - \setcounter{totalnumber}{6}  
 - \usepackage{caption, subcaption}
 - \usepackage{placeins}

@@

As shown in Figure  \ref{chat-pct}, groups vary substantially in how many trials they use the chat on, but many groups rarely use the chat. The horizontal lines correspond to 10% and 50% of the time, so games lower than the 10% line are using chat on fewer than 4 (out of 40) trials. There's a lot of not chatting in the access-to-chat condition.

```{r chat-pct, fig.width=10, fig.height=3, fig.cap="In chat conditions, how many trials each game choose to use the chat at all. Within each experiment, games are arranged in frequency of chat use. Horizontal lines denote 10% and 50% use of chat. \\label{chat-pct}"}

@@

game_chat |>
  filter(chat_cond == "chat") |>
  mutate(is.chat = ifelse(words > 0, 1, 0)) %>%
  group_by(gameId, expt) %>%
  summarize(pct_chat = mean(is.chat)) |>
  arrange(expt, pct_chat) |>
  group_by(expt) |>
  mutate(x = 1 / n() * row_number()) |>
  ggplot(aes(x = x, y = pct_chat)) +
  geom_point() +
  facet_grid(.~expt) + 
  geom_hline(yintercept = .1) +
  geom_hline(yintercept = .5) +
  labs(x="Fraction games", y="Fraction trials with chat messages")+
    scale_x_continuous(breaks=c(0, .25, .5, .75, 1), labels=c("0", ".25", ".5", ".75", "1"))+
      scale_y_continuous(breaks=c(0, .25, .5, .75, 1), labels=c("0", ".25", ".5", ".75", "1"))
```

@@

```{r}
msum_loc="manuscript/model_files/summary"

payout_mega <- read_rds(here(msum_loc, "payout_mega.rds"))

payout_combo <- read_rds(here(msum_loc,"payout_mega_combo.rds"))
payout_3 <- read_rds(here(msum_loc, "payout_combo_3.rds"))
payout_4 <- read_rds(here(msum_loc, "payout_combo_4.rds"))
payout_5 <- read_rds(here(msum_loc, "payout_combo_5.rds"))

# 
bos_aligned <- read_rds(here(msum_loc, "BoS_aligned_mega.rds"))
bos_3 <- read_rds(here(msum_loc, "BoS_aligned_3.rds"))
bos_4 <- read_rds(here(msum_loc, "BoS_aligned_4.rds"))
bos_5 <- read_rds(here(msum_loc, "BoS_aligned_5.rds"))

PD_cooperate <- read_rds(here(msum_loc, "PD_cooperate_mega.rds"))
PD_unneven <- read_rds(here(msum_loc, "PD_unneven_mega.rds"))

@@

The first way we assess whether access to chat makes a difference on game performance is to look at the mean number of points received per trial. Because BoS and PD have different point structures, we compare within game type.

```{r points, fig.cap="Rewards achieved per trial depending on experiment, game condition, and chat condition. \\label{points}"}

pal = c("PD_easy"="#1B9E77FF","PD_normal"= "#D95F02FF", "PD_hard"="#7570B3FF", "BoS_normal"= "#E7298AFF", "BoS_spike"="#66A61EFF")
bonuses <- all_rounds %>%
  group_by(playerId, gameId, chat_cond, cond, gametype, expt) %>%
  summarize(payoff = mean(payoff)) |> 
  mutate( combo_type=str_c(gametype, "_", cond) |> factor(levels=c("PD_easy", "PD_normal", "PD_hard", "BoS_normal", "BoS_spike")))

ggplot(bonuses, aes(x = chat_cond, y = payoff, color = combo_type, group = cond)) +
  geom_point(position = position_jitterdodge(jitter.width = .1, dodge.width = .7), alpha = .3) +
  facet_grid(gametype~expt) +
  scale_color_manual(values=pal) +
  labs(y = "mean_reward", x = "condition") +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .7), size = .3, color = "black") +
    stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .7), size = .7, color = "black", geom="line") +
guides(color = guide_legend(override.aes = list(alpha = 1, size=3)))+
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0, 20)) +
  labs(y = "Mean reward / trial", ) +
  theme(
    legend.title=element_blank(),
    axis.title.x=element_blank()
  )
```

In a model of payouts by coarse conditions (BoS v PD), there is an interaction between condition and chat condition where chat condition leads to higher payout for BoS (`r stats_text(payout_mega, 4)`). Using finer grained measures of condition, and treating "easy PD" as baseline, there are interactions between chat and both normal BoS (`r stats_text(payout_combo, 7)`) and spike BoS (`r stats_text(payout_combo, 8)`). The same directionality occurs in individual experiments, but with wide credible intervals. In Experiment 3, the chat interaction with normal BoS is `r stats(payout_3, 6)` and with spike BoS is `r stats(payout_3, 7)`. In Experiment 4, the interaction between chat and normal BoS is `r stats(payout_4, 4)`. In Experiment 5, the chat interaction with normal BoS is `r stats(payout_5, 6)` and with spike BoS is `r stats(payout_5, 7)`. Full model results are in the Appendix.

@@

@@

```{r}
outcome <- all_rounds %>%
  select(gameId, repNum, response, chat_cond, gametype, cond, role) |>
  pivot_wider(names_from = role, values_from = response) %>%
  mutate(outcome = str_c(p1, p2)) %>%
  left_join(game_chat) |> 
   mutate( combo_type=str_c(gametype, "_", cond) |> factor(levels=c("PD_easy", "PD_normal", "PD_hard", "BoS_normal", "BoS_spike")))

For BoS, the desirable outcome is to both choose the same box. As can be seen in Figure \@ref(fig-bosquad), this good outcome is achieved at above change levels when people can chat and usually at chance levels when people cannot. Across experiments, the chat condition has a higher rate of good (aligned) outcomes (`r stats_text(bos_aligned, 3)`). The estimates are less certain for individual experiments. In experiment 3 the effect was `r stats_text(bos_3, 3)`; in experiment 4, `r stats_text(bos_4,2)`; and in experiment 5, `r stats_text(bos_5,3)`.

```{r bosquad, fig.cap="For BoS trials, how frequently pairs made coordinated selections that recieved rewards. \\label{bosquad} ", fig.height=3, fig.width=8}
outcome %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = str_c(chat_cond), y = outcome_parity, color=combo_type, group=combo_type)) +
  facet_grid(~ expt) +
  scale_color_manual(values=pal)+
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  labs(y = "Good outcome?", x="Chat condition", color="")+
  theme(legend.position="bottom")
```
We again note that the off-diagonal rewards were misdisplayed for one of the players in experiments 1 and 2 for PD. On most trials "normal_PD" will pattern with easy PD.

@@

```{r , fig.cap="For PD trials, how frequently pairs made selections that results in each type of outcome. AA is the cooperate outcome, AB/BA is mixed -- one cooperate, one defect, BB is both defect. \\label{pdquad} ", fig.height=5, fig.width=8}
outcome %>%
  filter(gametype == "PD") %>%
  mutate(AA = ifelse(outcome %in% c("AA"), 1, 0)) %>%
  mutate(AB_BA= ifelse(outcome %in% c("AB", "BA"), 1, 0)) %>%
  mutate(BB = ifelse(outcome %in% c("BB"), 1, 0)) %>%
  pivot_longer(c("AA", "AB_BA", "BB"), names_to="outcome_type", values_to="selected") |> 
  ggplot(aes(x = chat_cond, y = selected, color=combo_type)) +
  facet_grid(outcome_type ~ expt) +
    scale_color_manual(values=pal)+

  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  labs(y="fraction selected", x="Chat condition", color="")+
  theme(legend.position="bottom")
```
From Figure TODO, we see that the preferred outcome is AA which is welfare-maximizing in easy PD. This is not what game theory predicts for rational actors, instead, participants are being nice and cooperative with each other.

Overall, for PD, the preference was to align on the fair-but-high square, regardless of whether chat was available. In the hard PD, AA selections were less common (`r stats_text(PD_cooperate, 2)`) and AA selections were slightly lowered still by a hard PD x chat interaction (`r stats_text(PD_cooperate, 5)`). The interaction effect had a credible interval overlapping 0 when experiments 3 or 5 were analyzed individually.  In parallel, there was a hard PD x chat interaction increasing the rate of AB/BA selections (`r stats_text(PD_unneven, 5)`), consistent with the idea that this is the best outcome for hard PD, but requires more complex coordination to achieve.

Overall, the results based on condition assignment are consistent with the prediction that language would be more beneficial for BoS than for PD. However, conditional differences are small and often noisy (presumably due to low update of using language), and so the experiments do not provide compelling evidence.

## Uptake results

To confirm that language use is, indeed useful, we also provide visualizations looking at both the option and choice to use language on a trial by trial basis. Note that a pairs' choice to use language is not assigned and thus this exploratory analysis could be confounded.

```{r}
outcome_chatted <- outcome |> 
  mutate(chatted = case_when(
    chat_cond =="nochat" ~ "nochat",
    is_chat == 1 ~ "chatted",
    is_chat == 0 ~ "quiet"
  ) |> factor(levels=c("chatted", "quiet", "nochat")))
```

@@

```{r, fig.cap="For BoS trials, how frequently pairs made coordinated selections that recieved rewards as a function of whether chat was available or used that trial. \\label{bosquad-uptake} ", fig.height=3, fig.width=8}
outcome_chatted %>%
  filter(gametype == "BoS") %>%
  mutate(outcome_parity = ifelse(outcome %in% c("AA", "BB"), 1, 0)) %>%
  ggplot(aes(x = chatted, y = outcome_parity, color=combo_type, group=combo_type)) +
  facet_grid(~ expt) +
  scale_color_manual(values=pal)+
  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  labs(y = "Good outcome?", x="Chat usage", color="")+
  theme(legend.position="bottom")
```

Near chance if you can't talk, above chance if you could but don't, far above chance if you do. This pattern suggests that some of the benefits to chatting spill over onto other trials, perhaps via setting up conventions for determining what strategy to use to break the symmetry.

```{r, fig.cap="For PD trials, how frequently pairs made selections that results in each type of outcome as a function of chat availability and usage on each trial. AA is the cooperate outcome, AB/BA is mixed -- one cooperate, one defect, BB is both defect. \\label{pdquad} ", fig.height=5, fig.width=8}
outcome_chatted %>%
  filter(gametype == "PD") %>%
  mutate(AA = ifelse(outcome %in% c("AA"), 1, 0)) %>%
  mutate(AB_BA= ifelse(outcome %in% c("AB", "BA"), 1, 0)) %>%
  mutate(BB = ifelse(outcome %in% c("BB"), 1, 0)) %>%
  pivot_longer(c("AA", "AB_BA", "BB"), names_to="outcome_type", values_to="selected") |> 
  ggplot(aes(x = chatted, y = selected, color=combo_type)) +
  facet_grid(outcome_type ~ expt) +
    scale_color_manual(values=pal)+

  stat_summary(fun.data = "mean_cl_boot", position=position_dodge(width=.3)) +
  geom_hline(aes(yintercept = .5)) +
  theme(legend.position="bottom")
```
Noteably, the selection of AB/BA options in hard PD is primarly being driven by trials where people did communicate with each other.

Another question is whether the amount of communication on a trial matters. Are more words more helpful?

```{r}
game_chat |> filter(words > 0) |> 
   mutate( combo_type=str_c(gametype, "_", cond) |> factor(levels=c("PD_easy", "PD_normal", "PD_hard", "BoS_normal", "BoS_spike"))) |> 
  ggplot(aes(x = words, y = mean_payout, color=combo_type)) +
  scale_x_log10(breaks=c(1,5,25))+
  scale_color_manual(values=pal)+
  geom_point(alpha=.1) +
  facet_grid(gametype ~ expt) +
  geom_smooth(method = "lm", show.legend=F)+
  guides(color = guide_legend(override.aes = list(alpha = 1, size=3)))+
  theme(legend.position="bottom")+
  labs(color="", x="Words in a trial", y="Payout")

#TODO label in not log space 
```

@@

A qualitative perusal of the data suggests various strategies. On a trial by trial reference basis, cells can be identified by color or color combination, sometimes even reducing to single letter (such as r for red). The numeric values can also be used to identify.

Between the chat logs and the post game strategy notes, a number of strategies were identified. For example, some participants coordinated using the running score totals choosing the option that favored the person with a lower current score. Other identified strategies included fairness, being generous to the other person, or generally being fair but occassionally trying to take advantage and get a higher payout.

@@

3. The repeated nature of the game and extra information such as running score allowed for coordinating in ways not available in the base single-shot games.

One of the hopes was to see if using language on some perhaps early trials would set up conventions or patterns that could then be successfully executed on later with no or little additional language.

@@

## Payout models

These models all predict bonus score as an outcome in terms of either the coarse grained conditions (PD v BoS) or the finer grained conditions, all with interactions with assigned chat condition.

Priors were normal(5,2) for intercept and normal(0,2) for beta and sd. Model were run both over all the experiments, and individually on each of experiments 3-5. Due to errors, we do not run separate models on 1 and 2.

```{r}
do_table("payout_mega.rds", cap="Mega-analytic model predicting bonus based on coarse game type and chat condition.")
do_table("payout_mega_combo.rds", cap="Mega-analytic model predicting bonus based on fine-grained game type and chat condition.")
do_table("payout_combo_3.rds", cap="Model predicting bonus for experiment 3. ")
do_table("payout_combo_4.rds", cap="Model predicting bonus for experiment 4. ")
do_table("payout_combo_5.rds", cap="Model predicting bonus for experiment 5. ")

@@

\FloatBarrier

## BoS models

Abstracting over the randomly selected payout values, we also examined each class of game by where the selections fell in the 2x2 grid. For BoS, due to symmetry, outcomes are either aligned (positive outcome) or misaligned (0 value outcome). We can look at the effect of chat on getting aligned outcomes.

These were logistic models with a prior of normal(0,1) for beta and sd.

```{r}
do_table("BoS_aligned_mega.rds", cap="Mega-analytic model predicting occurrance of aligned outcome in BoS. ")
do_table("BoS_aligned_3.rds", cap="Model predicting aligned BoS outcome for experiment 3. ")
do_table("BoS_aligned_4.rds", cap="Model predicting aligned BoS outcome for experiment 4. ")
do_table("BoS_aligned_5.rds", cap="Model predicting aligned BoS outcome for experiment 5. ")

## PD models

For PD, there are 3 different outcomes. The AA cooperative outcome, the AB/BA outcome, and the Nash equilibrium BB outcome. We ran separate logistic models for the occurrence of each outcome, although note that model results are inherently interdependent because each trial has exactly 1 of these outcomes.

These were logistic models with a prior of normal(0,1) for beta and sd.

```{r}
do_table("PD_cooperate_mega.rds", cap="Mega-analytic model predicting AA outcomes in PD trials")
do_table("PD_cooperate_3.rds", cap="Model predicting AA outcomes in PD trials for experiment 3.")
do_table("PD_cooperate_4.rds", cap="Model predicting AA outcomes in PD trials for experiment 4.")
do_table("PD_cooperate_5.rds", cap="Model predicting AA outcomes in PD trials for experiment 5.")

do_table("PD_unneven_mega.rds", cap="Mega-analytic model predicting AB/BA outcomes in PD trials. ")
do_table("PD_unneven_3.rds", cap="Model predicting AB/BA outcomes in PD trials for experiment 3.")
do_table("PD_unneven_4.rds", cap="Model predicting AB/BA outcomes in PD trials for experiment 4.")
do_table("PD_unneven_5.rds", cap="Model predicting AB/BA outcomes in PD trials for experiment 5.")

do_table("PD_nash_mega.rds", cap="Mega-analytic model predicting BB (Nash equilibrium) outcomes in PD trials.")
do_table("PD_nash_3.rds", cap="Model predicting BB outcomes in PD trials for experiment 3.")
do_table("PD_nash_4.rds", cap="Model predicting BB outcomes in PD trials for experiment 4.")
do_table("PD_nash_5.rds", cap="Model predicting BB outcomes in PD trials for experiment 5.")
